%!TEX root = main.tex

% =====================
% PREFACE
% ---------------------
% =====================
\chapter*{Preface}

This MSc.~thesis has been prepared by Aish Fenton at the University of Auckland, Department of Computer Science. It has been supervised by Dr.~Michael Dinneen. The work undertaken in this thesis has grown out of a research project sponsored by New Zealand Trade and Enterprise (NZTE) for the company vWorkApp Inc.~to research vehicle route optimisation for use within their software product.

\section*{Acknowledgements}

I'd like to thank my partner in crime, Anna Jobsis, for her encouragement, cajoling, threatening, bribing, doing my dishes, proof reading, guilt tripping, grammar policing, comforting\ldots\ doing whatever it took to help me get it done. Also, I'd like to thank my mum and dad for their encouragement and for always making me feel like higher education was within my reach.

Thanks to Steve Taylor and Steve Harding for holding down the fort at work while I disappeared to do this thesis. And likewise thanks to my team, Jono, Rash, Elena, Bob, Marcus, Yuri, and Robin for being on the ball (as always) despite my absence. Thanks to Brendon Petrich and vWorkApp Inc.~for providing me with time off work and being supportive of me undertaking this. 

And lastly, I especially owe Dr.~Michael Dinneen, my supervisor, a big thank you for persevering with me even though he must have doubted that I'd ever finish.

% =====================
% Abstract
% =====================
\cleardoublepage

\begin{center}
   \begin{minipage}{300pt}
   \begin{center}
      \textbf{Abstract}
   \end{center}
   In this thesis we present a new algorithm for the Vehicle Routing Problem called the Enhanced Bees Algorithm. It is adapted from a fairly recent algorithm, the Bees Algorithm, which was developed for continuous optimisation problems. We show that the results obtained by the Enhanced Bees Algorithm are competitive with the best meta-heuristics available for the Vehicle Routing Problem---it is able to achieve results that are within 0.5\% of the optimal solution on a commonly used set of test instances. We show that the algorithm has good runtime performance, producing results within 2\% of the optimal solution within 60 seconds, making it suitable for use within real world dispatch scenarios. Additionally, we provide a short history of well known results from the literature along with a detailed description of the foundational methods developed to solve the Vehicle Routing Problem.
   \end{minipage}
\end{center}

\cleardoublepage
